---
title: "simweights: similarity-based weighting for TCR differential abundance analysis"
author: "Radu Diaconescu & Thomas Smits"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TCRsimilift-Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(TCRsimilift)
```

# Introduction

The TCRsimilift package is a lightweight R implementation of the data pre-processing pipeline described in [Buytenhuijs et al. (2024). "Differential T cell receptor gene expression analysis using the Wilcoxon test with similarity-based weighting"](https://www.biorxiv.org/content/10.1101/2025.03.28.645951v1).

The package pre-processes immunological sequencing data from T-cell receptor sequencing data by adjusting read counts based on the expression levels of highly similar TCRs. This improves true positive rates in differential abundance (DA) analysis.

# Data Input Format

TCRsimilift expects a dataframe formatted according to [AIRR Standards 1.6](https://docs.airr-community.org/en/stable/datarep/rearrangements.html). The following columns are required:

| Column | Description |
|---|---|
| `sequence_id` | Unique identifier for each TCR sequence |
| `cdr3_aa` | Amino acid sequence of the CDR3 region |
| `consensus_count` | Read count for this sequence in this sample |
| `v_call` | V gene with allele (e.g. `TRBV12-3*01`) |
| `j_call` | J gene with allele (e.g. `TRBJ1-2*01`) |
| `sample_processing_id` | Sample identifier |

The package includes a minisubset for quick testing, and a full dataset of 22,000 rows from a mice experiment:

```{r}
head(mouse_PBSvTCZ_data_minisubset[, c("sequence_id", "cdr3_aa", "consensus_count",
                                        "v_call", "j_call", "sample_processing_id")])
```

```{r}
unique(mouse_PBSvTCZ_data_minisubset$sample_processing_id)
```

# Running the Pipeline

## adjust_counts()

The main entry point is `adjust_counts()`. It validates the input, aggregates counts per sequence per sample, computes similarity matrices for sequences sharing the same CDR3 length and V-J gene pair, and returns a dataframe with similarity-adjusted counts:

```{r}
results <- adjust_counts(mouse_PBSvTCZ_data_minisubset)
```

Key parameters:

- `sim_method`: `"HAMMING"` (default), `"BLOSUM"`, or custom function — see [Similarity Methods](#similarity-methods) below.
- `cutoff`: similarity score threshold (default `0.8`) — see [Cutoff Parameter](#cutoff-parameter) below.

# Outputs

`adjust_counts()` returns a dataframe with one row per sequence per sample. The key output column is `wrc` (weighted read count), the weighted `consensus_count`:

```{r}
head(results[, c("sequence_id", "sample_processing_id", "consensus_count", "wrc")])
```

A sequence keeps its original count if it has no neighbours above the similarity cutoff. Otherwise its count is adjusted toward the weighted average of its neighbourhood.

## as_counts_matrix()

To use the counts in a DA workflow, convert the output to a wide count matrix with `as_counts_matrix()`. Rows are sequences, columns are samples. By default it returns the similarity-adjusted counts (`wrc`), With `weighted = FALSE` it gives the original counts:

```{r}
count_matrix <- as_counts_matrix(results)
count_matrix_unweighted <- as_counts_matrix(results, weighted = FALSE)
count_matrix[1:4, ]
```

Setting `doFilter = TRUE` removes sequences that do not appear with at least one count in two or more samples, reducing noise before DA analysis:

```{r}
count_matrix_filtered <- as_counts_matrix(results, doFilter = TRUE)
dim(count_matrix_filtered)
```

# Downstream Analysis

## Wilcoxon Test (via edgeR normalisation)

The following workflow normalises the count matrix with edgeR's TMM method and applies a Wilcoxon test per sequence. This is the approach described in the accompanying pre-print.

```{r, eval=FALSE}
results <- adjust_counts(mouse_PBSvTCZ_data)
count_matrix <- as_counts_matrix(results, doFilter = TRUE)

library(edgeR)

class(count_matrix) <- "numeric"
sample_size <- 2
condition <- factor(c(rep("treatment", sample_size), rep("non-treatment", sample_size)))

y <- DGEList(counts = count_matrix, group = condition)
y <- calcNormFactors(y, method = "TMM")
count_norm <- as.data.frame(cpm(y))

pvalues <- sapply(1:nrow(count_norm), function(i) {
  data <- cbind.data.frame(gene = as.numeric(t(count_norm[i, ])), condition)
  wilcox.test(gene ~ condition, data)$p.value
})
pvalues <- data.frame(pvalues, row.names = rownames(count_norm))
pvalues$fdr <- p.adjust(pvalues$pvalues, method = "fdr")

conditionsLevel <- levels(condition)
dataCon1 <- count_norm[, which(condition == conditionsLevel[1])]
dataCon2 <- count_norm[, which(condition == conditionsLevel[2])]
dataCon2 <- dataCon2[rownames(dataCon1), ]
pvalues <- pvalues[rownames(dataCon1), ]

foldChanges <- log2((rowMeans(dataCon2) + 1) / (rowMeans(dataCon1) + 1))
results_da <- data.frame(
  log2foldChange = foldChanges,
  pValues = pvalues$pvalues,
  FDR = pvalues$fdr,
  row.names = rownames(dataCon1)
)
results_da <- na.omit(results_da)
```

## DESeq2

DESeq2 requires integer counts. Round the `wrc` values before passing them in:

```{r, eval=FALSE}
results <- adjust_counts(mouse_PBSvTCZ_data)
count_matrix <- as_counts_matrix(results, doFilter = TRUE)

library(DESeq2)

count_matrix_int <- round(count_matrix)
col_data <- data.frame(
  condition = factor(c(rep("treatment", 2), rep("control", 2))),
  row.names = colnames(count_matrix_int)
)

dds <- DESeqDataSetFromMatrix(countData = count_matrix_int,
                               colData = col_data,
                               design = ~ condition)
dds <- DESeq(dds)
res <- results(dds, contrast = c("condition", "treatment", "control"))
res_df <- as.data.frame(res[order(res$padj), ])
```

# Parameter Details

## Similarity Methods

The `sim_method` parameter controls how sequence similarity is computed within each CDR3 length / V-J gene group.

**`"HAMMING"` (default):** normalised Hamming distance between equal-length CDR3 amino acid sequences. A similarity of 1 means the sequences are identical; each mismatched position reduces the score by `1 / length`. Fast and suitable for most use cases.

**`"BLOSUM"`:** pairwise alignment scores using the BLOSUM62 substitution matrix, normalised to [0, 1]. Accounts for amino acid biochemical properties — conservative substitutions (e.g. Leu → Ile) receive a higher similarity than radical ones (e.g. Leu → Asp). More biologically informed than Hamming, but slower.

```{r, eval=FALSE}
results_hamming <- adjust_counts(mouse_PBSvTCZ_data_minisubset, sim_method = "HAMMING")
results_blosum <- adjust_counts(mouse_PBSvTCZ_data_minisubset, sim_method = "BLOSUM")
```

**Custom function:** you can also pass any function that takes a dataframe of sequences and returns a square similarity matrix with values in [0, 1]. The function is called once per CDR3 length / V-J group:

```{r, eval=FALSE}
my_sim <- function(seq_df) {
  matrix(0.9, nrow = nrow(seq_df), ncol = nrow(seq_df))
}
results_custom <- adjust_counts(mouse_PBSvTCZ_data_minisubset, sim_method = my_sim)
```

## Cutoff Parameter

The `cutoff` parameter (default `0.8`) is the minimum similarity score for two sequences to be treated as neighbours. Sequences below this threshold do not affect each other's adjusted counts.

- A **higher** cutoff (e.g. `0.9`) is more stringent: only nearly identical sequences are pooled together.
- A **lower** cutoff (e.g. `0.7`) allows more distant sequences to contribute, increasing smoothing but risking blending of biologically distinct clonotypes.

```{r}
results_strict <- adjust_counts(mouse_PBSvTCZ_data_minisubset, cutoff = 0.9)
results_lenient <- adjust_counts(mouse_PBSvTCZ_data_minisubset, cutoff = 0.7)
```
